Biometric analysis of infant mortality, Scania 1710--1812

### 2023-03-06 scania.Rmd
first commit

---
title: "A hazards approach to the biometric analysis of infant mortality"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
- \subsectionfont{\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested. In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age. Instead, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying. Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
---

@@

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

```{r first, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, Bourgeois-Pichat method."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
par(oldpar)
```

The cumulative numbers of death are plotted against the
corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
after 28 days (68 on the $g$ scale) is excellent for both time periods.

# The hazard-based procedure

Instead of using the total number of births throughout in the denominator as in  the
Bourgeois-Pichat procedure, we suggest 
using the present risk-set size, that is, the the total number of infants still alive 
immediately prior to the death age under investigation. It is easily achieved by the use of 
the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
an illustration.
```{r risksets, echo = FALSE, results = 'asis'}
source("R/tbl.R")
lt1 <- with(r1, cbind(ginv(risktimes), round(risktimes, 1), size, n.events))
lt1 <- data.frame(lt1)
names(lt1) <- c("Day", "g(Day)", "Population", "Deaths")
tbl(lt1[1:7, ], caption = "Life table, 1861-1890.")
```
It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
on average at the age of six hours (a quarter of a day), and so on.

The plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).

```{r second, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, hazards method."}
source("R/drawBB.R")
oldpar <- par(mfrow = c(1,2))
##
##r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
##r1 <- eha::risksets(r1, members = FALSE)
drawBB(r1, main = "1861-1890", ylim = c(0, 0.115))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
##r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
##r3 <- eha::risksets(r3, members = FALSE)
drawBB(r3, main = "1921-1950", ylim = c(0, 0.115))
par(oldpar)
```

See Figure \@ref(fig:both) for a comparison of the two curves.

```{r both, fig.cap = "Bourgeois-Pichat and 'hazards method' plots."}
oldpar = par(mfrow = c(1, 2), cex = 0.7, las = 1)
plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
     main = "1861-1890", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r1$risktimes, cumsum(r1$n.events / r1$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.08, "Hazard method", col = "red")
text(200, 0.02,"Bourgeois-Pichat", col = "blue")
##
plot(r3$risktimes, cumsum(r3$n.events) / r3$size[1], type = "l", col = "blue",
     main = "1921-1950", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r3$risktimes, cumsum(r3$n.events / r3$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.05, "Hazard method", col = "red")
text(200, 0.015,"Bourgeois-Pichat", col = "blue")
##
##
par(oldpar)
```

The difference between the two curves increases as infant mortality increases, 
but both methods show an excellent fit to linearity.

# Theoretical considerations

We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
to follow-up before one year of age) are present, and no left truncations (infants under observation 
only from an age later than birth). The hazards approach, on the other hand, allows
left truncation and right censoring as long as they are *non-informative* in the usual sense. Often, though,
this not very important, since new-born babies tend to be geographically stationary
for their first year of life.

Throughout the rest of the paper, everything happens on the $g$ time scale.

## Post-neonatal mortality

 We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).

In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.

```{r postneo, fig.cap = "Exponential fits to postneonatal data, Västerbotten."}
oldpar <- par(mfrow = c(1, 2), cex = 0.7)
post <- age.window(ginfant, c(g(28), g(365)))
p1 <- post[post$period == "1861-1890", ]
p3 <- post[post$period == "1921-1950", ]
fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue")
whaz1 <- hazards(wfit1)
with(whaz1, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
##
fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
whaz <- hazards(wfit3)
plot(fit3, main = "1921-1950", printLegend = FALSE, xlab = "g(Day)")
with(whaz, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
par(oldpar)
```

The *Exponential* fit is excellent, with a slight edge for the early period data, 
where post-neonatal mortality is high. However, one question pops up: If the exponential 
assumption is in fact true, will it also hold for subgroups of the data?

The theoretical ansver is "No", because a mixture of two exponential populations with
different mortality rates cannot itself be exponential. A heuristic argument for that: Assume two equally-sized 
groups at "birth", but with different levels of mortality. As time passes, there will be more deaths 
in the high-mortality group, and that means that the proportion survivors in the low-level group
will increase. Therefore, the population-level mortality will seem to decrease over time, 
and not being constant, as the exponentiality prescribes.

We can see what happens when the two periods in Figure \@ref(fig:postneo) are joined, see Figure
\@ref(fig:joined).

```{r joined, fig.cap = "Exponential fit for two periods, Västerbotten."}
oldpar = par(cex = 0.7)
p13 <- post[post$period != "1891-1920", ]
p13$period <- factor(p13$period)
fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
plot(fit13, main = "1861-1890 and 1921-1950", xlab = "g(Days)", printLegend = FALSE)
whaz13 <- hazards(wfit13)
with(whaz13, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
#
par(oldpar)
```

Not what we expected: The fit seems to be extremely good in both figures! However,
the rate is close to zero, and so differences are small and hard to notice. And for practical use, they are not interesting.

What about the original Bourgeois-Pichat model, is it sensitive to taking subsets?
Let us see.

```{r letse, fig.cap = "The period 1861-1890, Bourgeouis-Pichat method by sex."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(post[post$sex == "girl", ], eha::Surv(enter, exit, event))
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "Girls", ylim = c(0, 0.055))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
r3 <- with(post[post$sex == "boy", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "Boys", ylim = c(0, 0.055))
par(oldpar)
```

In Figure \@ref(fig:letse) we notice a weak tendency towards convexity for boys.
Not a big deal.

## Exogeneous mortality

We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase. @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant, 
19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
For instance, for $p = 0.2$ the fraction in question is 21.5 percent.

```{r exog, fig.height = 8, fig.cap = "Densities for uniform and exponential distributions for varying death probability p."}
source("R/denhaz.R")
p <- c(0.01, 0.05, 0.1, 0.2)
oldpar <- par(mfrow = c(2, 2))
for (i in 1:4){
   rr <- denhaz(p[i], header = paste("p =", p[i]))
}
par(oldpar)
```

## Endogeneous mortality

In order to calculate the *endogeneous* infant mortality, a simple two-step procedure leads
to the goal.

1.   Estimate the post-neonatal mortality following the exponential route. It reduces to a
simple occurrence/exposure calculation: The total number of post-neonatal deaths $D$ is divided by the
total postneonatal exposure $E$ on the $g$ scale. So
\begin{equation*}
\hat{\lambda} = \frac{D}{E}
\end{equation*}
is the estimated hazard function (constant), and the cumulative hazard function is
\begin{equation*}
\hat{H}(t) = \hat{\lambda} t, \; t > 0
\end{equation*}

2.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
$\hat{A}(t)$ to get $\hat{E}(t)$.
\begin{equation}
\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(28).
\end{equation}
Note that subtraction and addition of competing risks are okay on the hazards scale, 
but not with probabilities, which is yet another argument in favor of the hazards approach.

Let us do it with the given data, and the first period 1861--1890.

### Step 1: Estimate post-neonatal (exogenous) mortality

Post-neonatal mortality is the same as exogenous mortality in the sense that the
hazard functions are the same on the post-neonatal age interval. So the first step
involves data left truncated at age 28 days (or at `r round(g(28))`) on the $g$ scale.
We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
The first lines of the data set are (Table \@ref(tab:firstlines))
```{r firstlines, results='asis'}

tbl(infant[1:5, ], caption = "First five rows of data frame.")
nro <- nrow(infant)
deat <- as.integer(sum(infant$event))
```
and the number of rows is `r nro`, and the number of deaths is `r deat`,
resulting in a crude death probability of `r round(1000 * deat / nro)` per thousand
live births.

So, our first step is to truncate and rescale. We show how it is done in **R**:

```{r retrunc, echo = TRUE}
postneo <- eha::age.window(infant, c(28, 365))
gpostneo <- eha::age.window(ginfant, c(g(28), g(365)))
print(postneo[1:5, ], row.names = FALSE) # print first five rows
print(gpostneo[1:5, ], row.names = FALSE) # first five rows after...
```

Now, the post-neonatal hazard function is *constant*, and its ML estimator is simply 
the *occurrence/exposure* rate, or in **R**,

```{r occexp, echo = TRUE}
D <- sum(gpostneo$event) # No. of deaths
E <- with(gpostneo, sum(exit - enter)) # exposure
(rate <- D/E)
```

The rate is a very small number as a consequence of the very small time unit 
(implying large total exposure time), originally *day*.

So we are done with the simple post-neonatal period.

### Step 2: Estimate total and endogenous mortality

Here we focus on neonatal period.

The estimation of the cumulative hazard function for the total neonatal mortality is
easily achieved by calling the function `hazards` in the *eha* package, see also Figure \@ref(fig:hazards).

```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
neo <- age.window(ginfant, c(0, g(28)))
par(las = 1)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
hneo <- hazards(fit, cum = TRUE)
x <- c(0, hneo[[1]][, 1])
y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
lines(x, 1000 * rate * x, col = "blue", lty = 2)
abline(h = 0, v = 0)
text(8, 15, "Total", col = "black")
text(8, 5, "Exogenous", col = "blue")
```

Next, take the difference between "Total" and "Exogenous" to get "Endogenous", Figure \@ref(fig:enof).

```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
yen <- y - 1000 * rate * x
par(las = 1)
plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36))
abline(h = 0, v = 0)
lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
lines(x, y, lty = 1)
text(8, 20, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
```

@@

## Västerbotten 1921--1950

```{r lowume}
##source("R/drawBP.R")
##source("R/drawBB.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[ume$birthdate > "1920-12-31", ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

The crude IMR in Västerbotten 1921--1950 was around `r crudprob` per thousand, a rather 
low figure in context. Let us perform the biometric analysis with these data, see 
Figure \@ref(fig:lowumee).

```{r lowumee, fig.cap = "Västerbotten 1921--1950."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(28), g(365)))
##rs.post <- risksets()
ume <- ginfant[ginfant$birthdate > "1920-12-31", ]
pf(ume)
```

There is a good hazards model fit, and we see that the endogenous mortality clearly dominates 
the early days of neonatal mortality, and almost vanishes towards the start of the post-neonatal period.

The B-P model fit is slightly worse, but not much to bother about.

## Västerbotten 1890--1920

```{r medume}
source("R/biometric.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[(ume$birthdate > "1890-12-31") & (ume$birthdate <= "1920-12-31"), ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

The crude IMR in Umeå 1891--1920 was around `r crudprob` per thousand, clearly higher than 
the later time period. Let us perform the biometric analysis with these data, see 
Figure \@ref(fig:bioume9120).

```{r bioume9120, fig.cap = "Västerbotten 1891--1920."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(28), g(365)))
##rs.post <- risksets()
ume$exit <- g(ume$exit)
ume$enter <- g(ume$enter)
pf(ume)
```

The dominance of endogenous mortality in early life is still clear. The hazards fit is still slightly better.

## Västerbotten 1861--1890

```{r highume}
source("R/biometric.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[(ume$birthdate <= "1890-12-31"), ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

The crude IMR in Umeå 1861--1890 was around `r crudprob` per thousand, highest of the three
time periods.
The biometric analysis is shown in 
Figure \@ref(fig:bioume6190).

```{r bioume6190, fig.cap = "Västerbotten 1861--1890."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(28), g(365)))
##rs.post <- risksets()
ume$exit <- g(ume$exit)
ume$enter <- g(ume$enter)
pf(ume)
```

The conclusion here is almost the same as  for the later time periods, good model
fit and endogenous dominance in the very early days of life. However, the B-P method seems to have a slight upper hand.

## Scania 1710--1800.

```{r scahigh, fig.cap = "Scania 1710--1800."}
sca1 <- readRDS("Rdata/sca1.rds")
sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
sca1$exit <- g(sca1$exit)
sca1$enter <- g(sca1$enter)
rs <- with(sca1, risksets(Surv(enter, exit, event), members = FALSE))
oldpar <- par(mfrow = c(1, 2))
drawBB(rs, main = "Hazards method")
drawBP(rs, main = "B-P method")
par(oldpar)
```

Very bad fits in both cases.  <!--- Let us look at the estimated cumulative hazards, total, 
exogenous and endgenous, see Figure \@ref(fig:escu).

```{r escu, fig.cap = "Scania 1710-1800.", eval = FALSE}
#source("R/biometric.R")
#biometric(sca1, graph = TRUE)
post <- age.window(sca1, c(g(28), 365))
lambda <- with(post, sum(event) / sum(exit - enter))
x <- c(0, rs$risktimes)
y <- x * lambda
hazo <- with(rs, c(0, n.events / size))
hazo <- cumsum(hazo)
plot(x, hazo, type = "s", col = "blue")
lines(x, y)
z <- hazo - y
lines(x, z, col = "red")
```

--->

## Skellefteå 1820--1835

An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
There is no room for endogeneous deaths at all, but otherwise a reasonably good fit with both 
methods.

```{r skum, fig.cap = "Skellefteå 1821--1838."}
library(skum)
births <- obs[obs$starttyp == 2 &
                 obs$foddat > as.Date("1820-12-31") &
                 obs$foddat <= as.Date("1838-12-31"), ]
births$event <- births$sluttyp == 2
births$enter <- 365 * births$enter
births$exit <- 365 * births$exit
addup <- births$exit <= 0
births$exit[addup] <- 0.25
infa <- age.window(births, c(0, 365))
infa$enter <- g(infa$enter)
infa$exit <- g(infa$exit)
pf(infa)
```

>>>>

---
title: "Biometric analysis of infant mortality, Scania 1710--1812"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
- \subsectionfont{\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is studied in connection to data from Scania in the very south of Sweden."
---

@@

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs. In addition, we give examples from eighteenth century Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/sca1.rds")
infant$period <- cut(infant$BirthDate, 
                   breaks = as.Date(c("1709-12-31", "1734-12-31", 
                                      "1759-12-31","1784-12-31",  "1812-12-12")),
                   labels = c("1710-1734", "1735-1759", "1760-1784", "1785-1812"))
infant <- infant[, c("BirthDate", "Sex", "enter", "exit", "event", "period", 
                "socst", "ICD10")]

ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```

@@

## Scania 1710--1820.

```{r scahigh, fig.cap = "Scania 1710--1800."}
source("R/drawBB.R")
source("R/drawBP.R")
sca1 <- infant
sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
sca1$exit <- g(sca1$exit)
sca1$enter <- g(sca1$enter)
rs <- with(sca1, risksets(Surv(enter, exit, event), members = FALSE))
oldpar <- par(mfrow = c(1, 2))
drawBB(rs, main = "Hazards method")
drawBP(rs, main = "B-P method")
par(oldpar)
```

Very bad fits in both cases. Let us look at subsets of the data.

### Hazards by season

We define *season* as four three-months periods, *winter = December + January + February*,
*spring = March + April + May*, and so on.

```{r seasons, fig.cap = "BB by birth season."}
sca1$month <- as.numeric(format(as.Date(sca1$BirthDate, format = "%Y-%m-%d"), "%m"))
oldpar <- par(mfrow = c(2, 2))
sca1$Q <- ""
sca1$Q[sca1$month %in% c(1, 2, 12)] <- "winter"
sca1$Q[sca1$month %in% c(3, 4, 05)] <- "spring"
sca1$Q[sca1$month %in% c(6, 7, 08)] <- "summer"
sca1$Q[sca1$month %in% c(9, 10, 11)] <- "fall"
sca1$Q <- as.factor(sca1$Q)

for (when in levels(sca1$Q)){
   xxx <- with(sca1[sca1$Q %in% when, ], Surv(enter, exit, event))
   rs <- risksets(xxx, members = FALSE)
   drawBB(rs, main = when)
}
par(oldpar)
```

### Hazards without high-mortality years

```{r highmort, fig.cap = "Years with not so high IMR."}
infant$year <- with(infant, format(BirthDate, "%Y"))
oldpar <- par(mfrow = c(1, 2), cex = 0.7)
years <- sort(unique(infant$year))

n <- length(years)
imr <- numeric(n)
i <- 0
for (yr in years){
   i <- i + 1 
   imr[i] <- with(infant[infant$year == yr, ], sum(event) / (sum(exit - enter) / 365))
}
plot(years, imr, type = "b", las = 1, main = "IMR")
abline(h = 0.6, lty = 2, col = "red")

badyears <- years[imr > 0.6]
goodinf <- infant[!(infant$year %in% badyears), ]
goodinf$exit <- g(goodinf$exit)
goodinf$enter <- g(goodinf$enter)
rs <-  with(goodinf, risksets(Surv(enter, exit, event), members = FALSE))
drawBB(rs, main = "IMR less than 60 percent")
par(oldpar)
```

Tried a couple of cut points, no important differences.

### Hazards by socst

```{r socst, fig.cap = "Upper and lower classes."}
infa <- infant
oldpar <- par(mfrow = c(1, 2))
infa$exit <- g(infa$exit)
xxx <- with(infa[infa$socst %in% c("tjänsteman", "präst", "adel"), ], 
            Surv(enter, exit, event))
rs <- risksets(xxx, members = FALSE)
drawBB(rs, main = "Upper")
###
xxx <- with(infa[!(infa$socst %in% c("tjänsteman", "präst", "adel")), ], 
            Surv(enter, exit, event))
rs <- risksets(xxx, members = FALSE)
drawBB(rs, main = "Lower")
par(oldpar)
```
